Connecting R with D3 for dynamic graphics, to explore multivariate data with tours

### 2019-03-03 paper-tourrd3.Rmd
revisions to make it easier to understand

---
title: Connecting R with D3 for dynamic graphics, to explore multivariate data with tours
author:
  - name: Michael Kipp
    affiliation: Monash University
    address:
    - Department of Econometrics and Business Statistics
    email:  mkipp271@gmail.com
  - name: Dianne Cook
    affiliation: Monash University
    address:
    - Department of Econometrics and Business Statistics
    email:  dicook@monash.edu 
  - name: Ursula Laa
    affiliation: Monash University
    address:
    - School of Physics and Astronomy
    email:  ursula.laa@monash.edu
bibliography: references.bib
link-citations: true
fig_caption: yes
abstract: >
  The tourr package in R has several algorithms and displays for showing multivariate data as a sequence of low-dimensional projections. It can display as a movie but has no capacity for interaction, such as stop/go, change tour type, drop/add variables. The tourrGui  package provides these sorts of controls, but the interface is programmed with the dated RGtk2 package. This work explores using custom messages to pass data from R to D3  for viewing, using the Shiny framework.  
preamble: >
  % Any extra latex you need in the preamble
output: rticles::rjournal_article
---

@@

@@

The \pkg{tourr} package [@tourr] is an R implementation of the tour algorithms discussed in @gt_pp_mc. It includes methods for geodesic interpolation and basis generation, as well as an implementation of the simulated annealing algorithm to optimise projection pursuit indices for the guided tour. The tour can be displayed directly in the R graphics device, for example:

```{r eval=FALSE, echo=TRUE}
library(tourr)
# quartz() # to display on a Mac
# X11() # For windows
# The Rstudio graphics device is not advised
animate_dist(flea[, 1:6], center = TRUE)
```

generating the animation, with some stills shown in Figure \ref{tour}.

@@

>>>>

---
title: Connecting R with D3 for dynamic graphics, to explore multivariate data with tours
author:
  - name: Michael Kipp
    affiliation: Monash University
    address:
    - Department of Econometrics and Business Statistics
    email:  mkipp271@gmail.com
  - name: Ursula Laa
    affiliation: Monash University
    address:
    - School of Physics and Astronomy
    email:  ursula.laa@monash.edu
  - name: Dianne Cook
    affiliation: Monash University
    address:
    - Department of Econometrics and Business Statistics
    email:  dicook@monash.edu 
bibliography: references.bib
link-citations: true
fig_caption: yes
abstract: >
  The tourr package in R has several algorithms and displays for showing multivariate data as a sequence of low-dimensional projections. It can display as a movie but has no capacity for interaction, such as stop/go, change tour type, drop/add variables. The tourrGui  package provides these sorts of controls, but the interface is programmed with the dated RGtk2 package. This work explores using custom messages to pass data from R to D3  for viewing, using the Shiny framework. This is an approach that can be generally used for creating all sorts of interactive graphics.  
preamble: >
  % Any extra latex you need in the preamble
output: rticles::rjournal_article
---

@@

Dd you know you can run *any javascript you like* in a Shiny application and you can pass *whatever you want including JSON* back and forth? This massively widens the scope of what you can do with Shiny, and generating a tour of multivariate data with this approach is a really good example of what is possible.

@@

The \pkg{tourr} package [@tourr] is an R implementation of the tour algorithms discussed in @gt_pp_mc. It includes methods for geodesic interpolation and basis generation, as well as an implementation of the simulated annealing algorithm to optimise projection pursuit indices for the guided tour. The tour can be displayed directly in the R graphics device, for example, the code below generates a 1D density tour. Figure \ref{tour} shows snapshots.

```{r eval=FALSE, echo=TRUE, tidy=TRUE}
library(tourr)
# quartz() # to display on a Mac;
# X11() # For windows;
# The Rstudio graphics device is not advised
animate_dist(flea[, 1:6], center = TRUE)
```

@@

## Try it

You can try the app yourself using this code:

```{r eval=FALSE, echo=TRUE, tidy=TRUE}
devtools::install_github("uschiLaa/tourrGUID3")
library(tourrGUID3)
launchApp(system.file("extdata", "geozoo.csv", package = "tourrGUID3"))
```

### 2019-03-03 paper-tourrd3.tex
revisions to make it easier to understand

% !TeX root = RJwrapper.tex
\title{Connecting R with D3 for dynamic graphics, to explore multivariate data
with tours}
\author{by Michael Kipp, Dianne Cook, Ursula Laa}

@@

\abstract{%
The tourr package in R has several algorithms and displays for showing
multivariate data as a sequence of low-dimensional projections. It can
display as a movie but has no capacity for interaction, such as stop/go,
change tour type, drop/add variables. The tourrGui package provides
these sorts of controls, but the interface is programmed with the dated
RGtk2 package. This work explores using custom messages to pass data
from R to D3 for viewing, using the Shiny framework.\\
}

@@

@@

The \pkg{tourr} package (Wickham et al.
\protect\hyperlink{ref-tourr}{2011}) is an R implementation of the tour
algorithms discussed in Cook et al.
(\protect\hyperlink{ref-gt_pp_mc}{2007}). It includes methods for
geodesic interpolation and basis generation, as well as an
implementation of the simulated annealing algorithm to optimise
projection pursuit indices for the guided tour. The tour can be
displayed directly in the R graphics device, for example:

\begin{Schunk}
\begin{Sinput}
library(tourr)
# quartz() # to display on a Mac
# X11() # For windows
# The Rstudio graphics device is not advised
animate_dist(flea[, 1:6], center = TRUE)
\end{Sinput}
\end{Schunk}

generating the animation, with some stills shown in Figure \ref{tour}.

@@

A tour path is a smooth sequence of projection matrices, \(p\times d\),
that when combined with a matrix of n data points, \(n\times p\), and a
rendering method, produces a steady stream of \(d\)-dimensional views of
the data. Each tour is initialised with the \texttt{new\_tour()} method,
which instantiates a tour object and takes as arguments the data \(X\),
the tour method, e.g.~\texttt{guided\_tour()}, and the starting basis.
Once initialised, a new target plane is chosen, and a series of steps
along a geodesic path from starting to target plane are generated by
interpolation.

@@

This code is from the observe environment from the \texttt{server.R}
file. It converts the matrix of projected data points to JSON format,
and sends it to JavaScript with the id data. The list entries of the
``message'' can parsed in D3 by its \texttt{data()} method,
e.g.~\texttt{data(message.d)} to access the projected data points, and
we can access each column through the column names assigned in the
original dataframe, and loop over all rows for rendering. All of the
code required to render the scatterplots and legends, along with
colours, is JavaScript code in the file \texttt{d3anim.js}. In
particular, the data from R is handled with the following code:

@@

>>>>

% !TeX root = RJwrapper.tex
\title{Connecting R with D3 for dynamic graphics, to explore multivariate data
with tours}
\author{by Michael Kipp, Ursula Laa, Dianne Cook}

@@

\abstract{%
The tourr package in R has several algorithms and displays for showing
multivariate data as a sequence of low-dimensional projections. It can
display as a movie but has no capacity for interaction, such as stop/go,
change tour type, drop/add variables. The tourrGui package provides
these sorts of controls, but the interface is programmed with the dated
RGtk2 package. This work explores using custom messages to pass data
from R to D3 for viewing, using the Shiny framework. This is an approach
that can be generally used for creating all sorts of interactive
graphics.\\
}

@@

Dd you know you can run \emph{any javascript you like} in a Shiny
application and you can pass \emph{whatever you want including JSON}
back and forth? This massively widens the scope of what you can do with
Shiny, and generating a tour of multivariate data with this approach is
a really good example of what is possible.

@@

The \pkg{tourr} package (Wickham et al.
\protect\hyperlink{ref-tourr}{2011}) is an R implementation of the tour
algorithms discussed in Cook et al.
(\protect\hyperlink{ref-gt_pp_mc}{2007}). It includes methods for
geodesic interpolation and basis generation, as well as an
implementation of the simulated annealing algorithm to optimise
projection pursuit indices for the guided tour. The tour can be
displayed directly in the R graphics device, for example, the code below
generates a 1D density tour. Figure \ref{tour} shows snapshots.

\begin{Schunk}
\begin{Sinput}
library(tourr)
# quartz() # to display on a Mac; X11() # For windows; The Rstudio graphics
# device is not advised
animate_dist(flea[, 1:6], center = TRUE)
\end{Sinput}
\end{Schunk}

@@

A tour path is a smooth sequence of projection matrices, \(p\times d\),
that when combined with a matrix of n data points, \(n\times p\), and a
rendering method, produces a steady stream of \(d\)-dimensional views of
the data. Each tour is initialised with the \texttt{new\_tour()} method,
which instantiates a tour object and takes as arguments the data \(X\),
the tour method, e.g. \texttt{guided\_tour()}, and the starting basis.
Once initialised, a new target plane is chosen, and a series of steps
along a geodesic path from starting to target plane are generated by
interpolation.

@@

This code is from the observe environment from the \texttt{server.R}
file. It converts the matrix of projected data points to JSON format,
and sends it to JavaScript with the id data. The list entries of the
``message'' can parsed in D3 by its \texttt{data()} method, e.g.
\texttt{data(message.d)} to access the projected data points, and we can
access each column through the column names assigned in the original
dataframe, and loop over all rows for rendering. All of the code
required to render the scatterplots and legends, along with colours, is
JavaScript code in the file \texttt{d3anim.js}. In particular, the data
from R is handled with the following code:

@@

\hypertarget{try-it}{%
\subsection{Try it}\label{try-it}}

You can try the app yourself using this code:

\begin{Schunk}
\begin{Sinput}
devtools::install_github("uschiLaa/tourrGUID3")
library(tourrGUID3)
launchApp(system.file("extdata", "geozoo.csv", package = "tourrGUID3"))
\end{Sinput}
\end{Schunk}
